---
title: "Module 1: Tabular Data"
subtitle: "Working with larger-than-RAM data using duckdbfs"
author: "ESPM 288"
format: html
---

## Introduction

In this module, we will explore high-performance workflows for tabular data. We will use `duckdbfs` to work with datasets that are larger than available RAM by leveraging DuckDB's streaming and remote file access capabilities.

## Case Study: Global Supply Chains

We will be working with [EXIOBASE 3.8.1](https://source.coop/youssef-harby/exiobase-3), a global Multi-Regional Input-Output (MRIO) database. This dataset tracks economic transactions between sectors and regions, along with their environmental impacts (emissions, resource use, etc.).

**Data description:**
- **Coverage**: 44 countries + 5 rest-of-world regions.
- **Timeframe**: 1995–2022.
- **Content**: Economic transactions (Z matrix), final demand (Y matrix), and environmental stressors (F matrix).
- **Format**: Cloud-optimized Parquet, partitioned by year and matrix type.

## Setup

```{r}
library(duckdbfs)
library(dplyr)
library(ggplot2)
```

## Exercise 1: connecting to remote data

We can open the entire dataset without downloading it using `open_dataset()`. The data is hosted on Source Cooperative. The `**` pattern allows recursive scanning of the partitioned parquet files.

```{r}
# Remote S3 path to EXIOBASE 3 (Source Cooperative)

duckdbfs::duckdb_secrets(
    key = "",
    secret = "",
    endpoint = "s3.amazonaws.com",
    region = "us-west-2"
)
s3_url <- "s3://us-west-2.opendata.source.coop/youssef-harby/exiobase-3/4588235/parquet/**"

# Open the dataset lazily
exio <- open_dataset(s3_url)

# View the schema (column names and types) without reading data
glimpse(exio)
```

## Exercise 2: Efficient Filtering

The dataset is large. We should filter *before* collecting any data into R.

```{r}
exio |>
    filter(year == 2022, region == "US") |>
    head() |> # view the first 6 rows
    collect()
```

> **Task**: Construct a query to find the top 5 sectors in the US by CO2 emissions in 2022. Remember to check the column names in `exio` to find the appropriate emissions flow.

```{r}
# Solution: Top 5 sectors in US by CO2 emissions (2022)

# Target the F_satellite matrix directly to access stressor column
exio_f_sat <- open_dataset(
    "s3://us-west-2.opendata.source.coop/youssef-harby/exiobase-3/4588235/parquet/year=2022/format=ixi/matrix=F_satellite/"
)

exio_f_sat |>
    filter(
        grepl("CO2", stressor)
    ) |>
    group_by(sector) |>
    summarize(total_co2 = sum(value, na.rm = TRUE), .groups = "drop") |>
    arrange(desc(total_co2)) |>
    head(5) |>
    collect()
```

## Exercise 3: CO2 Emissions Over Time by Country (Top 5 Emitters)

First, let's identify the top 5 CO2 emitting countries in 2022:

```{r}
# Find the top 5 emitting countries in 2022
top_emitters_2022 <- exio |>
    filter(year == 2022,
           matrix == "F_satellite",
           stressor %like% "%CO2%") |>
    group_by(region) |>
    summarise(total_co2 = sum(value, na.rm = TRUE)) |>
    filter(!grepl("^W", region)) |>  # Exclude rest-of-world regions
    arrange(desc(total_co2)) |>
    head(5) |>
    collect()

top_emitters_2022
```

Now, let's get the time series data for these top 5 countries across all years:

```{r}
# Get the list of top 5 countries
top_5_countries <- top_emitters_2022$region

# Query CO2 emissions for these countries across all years
co2_timeseries <- exio |>
    filter(matrix == "F_satellite",
           stressor %like% "%CO2%",
           region %in% top_5_countries) |>
    group_by(year, region) |>
    summarise(total_co2 = sum(value, na.rm = TRUE)) |>
    arrange(year, region) |>
    collect()

# View the data
head(co2_timeseries, 20)
```

Create a line plot showing CO2 emissions over time for the top 5 countries:

```{r}
# Create the time series plot
ggplot(co2_timeseries, aes(x = year, y = total_co2, color = region)) +
    geom_line(linewidth = 1.2) +
    geom_point(size = 2.5) +
    labs(
        title = "CO2 Emissions Over Time: Top 5 Emitting Countries",
        subtitle = "Based on EXIOBASE 3.8.1 data (1995-2022)",
        x = "Year",
        y = "Total CO2 Emissions",
        color = "Country/Region"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(face = "bold", size = 14),
        legend.position = "right",
        legend.title = element_text(face = "bold", size = 11),
        legend.text = element_text(size = 10),
        legend.background = element_rect(fill = "white", color = "gray80"),
        legend.key.size = unit(1, "cm")
    ) +
    scale_x_continuous(breaks = seq(1995, 2022, by = 5))
```

## Exercise 4: Countries with Largest CO2 Emission Reductions 
There are lots of stressors, how do I find the stressors related to CO2?

```{r}
# List all unique stressors containing 'CO2' to see available options
co2_stressors <- exio |>
    filter(matrix == "F_satellite") |>
    distinct(stressor) |>
    filter(grepl("CO2", stressor)) |>
    collect()
co2_stressors
```

Now, let's identify which countries have emitted the most CO2 emissions between 1995 and 2022:
```{r}
# Calculate total CO2 emissions for the top 5 countries across all years and countries
total_co2_by_country <- exio |>
    filter(matrix == "F_satellite") |>
    collect() |>
    filter(grepl("CO2|carbon Dioxide", stressor, ignore.case = TRUE))
```

```{r}
#Calculate which regions are the top 5 CO2 emitters
top_regional_emitters <- total_co2_by_country |>
    group_by(region) |>
    summarize(total_co2 = sum(value, na.rm = TRUE)) |>
    filter(!grepl("^W", region)) |>  # Exclude rest-of-world regions
    arrange(desc(total_co2)) |>
    head(5) |>
    collect()

# View the data first
print(top_regional_emitters)


```

```{r}

# Plot the top 5 regional emitters by total CO2 emissions from 1995 to 2022
print(ggplot(top_regional_emitters, aes(x = reorder(region, -total_co2), y = total_co2, fill = region)) +
    geom_bar(stat = "identity")
    labs(
        title = "Total CO2 Emissions by Country (1995-2022)",
        x = "Country/Region",
        y = "Total CO2 Emissions",
        fill = "Country/Region"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(face = "bold", size = 14),
        legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)
    ))
```  
