---
title: "Module 1: Tabular Data"
subtitle: "Working with larger-than-RAM data using duckdbfs"
author: "ESPM 288"
format: html
---

## Introduction

In this module, we will explore high-performance workflows for tabular data. We will use `duckdbfs` to work with datasets that are larger than available RAM by leveraging DuckDB's streaming and remote file access capabilities.

## Case Study: Global Supply Chains

We will be working with [EXIOBASE 3.8.1](https://source.coop/youssef-harby/exiobase-3), a global Multi-Regional Input-Output (MRIO) database. This dataset tracks economic transactions between sectors and regions, along with their environmental impacts (emissions, resource use, etc.).

**Data description:**
- **Coverage**: 44 countries + 5 rest-of-world regions.
- **Timeframe**: 1995–2022.
- **Content**: Economic transactions (Z matrix), final demand (Y matrix), and environmental stressors (F matrix).
- **Format**: Cloud-optimized Parquet, partitioned by year and matrix type.

## Setup

```{r}
library(duckdbfs)
library(dplyr)
library(ggplot2)
```

## Exercise 1: connecting to remote data

We can open the entire dataset without downloading it using `open_dataset()`. The data is hosted on Source Cooperative. The `**` pattern allows recursive scanning of the partitioned parquet files.

```{r}
# Remote S3 path to EXIOBASE 3 (Source Cooperative)

duckdbfs::duckdb_secrets(
    key = "",
    secret = "",
    endpoint = "s3.amazonaws.com",
    region = "us-west-2"
)

# Open the entire dataset with ** pattern for recursive scanning
s3_url <- "s3://us-west-2.opendata.source.coop/youssef-harby/exiobase-3/4588235/parquet/**"

# Open the dataset lazily
exio <- open_dataset(s3_url)

# View the schema (column names and types) without reading data
glimpse(exio)
```

## Exercise 2: Efficient Filtering

The dataset is large. We should filter *before* collecting any data into R.

```{r}
exio |>
    filter(year == 2022, region == "US") |>
    head() |> # view the first 6 rows
    collect()
```

> **Task**: Construct a query to find the top 5 sectors in the US by CO2 emissions in 2022. Remember to check the column names in `exio` to find the appropriate emissions flow.

```{r}
# First, check the column names in exio to find the right columns
glimpse(exio)
 
# Query CO2 emissions by sector for the US in 2022
co2_emissions <- exio |>
    filter(year == 2022, 
           region == "US",
           stressor %like% "%CO2%") |>
    group_by(sector) |>
    summarise(total_co2 = sum(value, na.rm = TRUE)) |>
    arrange(desc(total_co2)) |>
    head(5) |>
    collect()

co2_emissions
```

## Exercise 3: CO2 Emissions Over Time by Country (Top 5 Emitters)

First, let's identify the top 5 CO2 emitting countries in 2022:

```{r}
# Find the top 5 emitting countries in 2022
top_emitters_2022 <- exio |>
    filter(year == 2022,
           matrix == "F_satellite",
           stressor %like% "%CO2%") |>
    group_by(region) |>
    summarise(total_co2 = sum(value, na.rm = TRUE)) |>
    filter(!grepl("^W", region)) |>  # Exclude rest-of-world regions
    arrange(desc(total_co2)) |>
    head(5) |>
    collect()

top_emitters_2022
```

Now, let's get the time series data for these top 5 countries across all years:

```{r}
# Get the list of top 5 countries
top_5_countries <- top_emitters_2022$region

# Query CO2 emissions for these countries across all years
co2_timeseries <- exio |>
    filter(matrix == "F_satellite",
           stressor %like% "%CO2%",
           region %in% top_5_countries) |>
    group_by(year, region) |>
    summarise(total_co2 = sum(value, na.rm = TRUE)) |>
    arrange(year, region) |>
    collect()

# View the data
head(co2_timeseries, 20)
```

Create a line plot showing CO2 emissions over time for the top 5 countries:

```{r}
# Create the time series plot
ggplot(co2_timeseries, aes(x = year, y = total_co2, color = region)) +
    geom_line(linewidth = 1.2) +
    geom_point(size = 2.5) +
    labs(
        title = "CO2 Emissions Over Time: Top 5 Emitting Countries",
        subtitle = "Based on EXIOBASE 3.8.1 data (1995-2022)",
        x = "Year",
        y = "Total CO2 Emissions",
        color = "Country/Region"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(face = "bold", size = 14),
        legend.position = "right",
        legend.title = element_text(face = "bold", size = 11),
        legend.text = element_text(size = 10),
        legend.background = element_rect(fill = "white", color = "gray80"),
        legend.key.size = unit(1, "cm")
    ) +
    scale_x_continuous(breaks = seq(1995, 2022, by = 5))
```

## Exercise 4: Countries with Largest CO2 Emission Reductions 
This is just me trying something for fun to see what I can do here!!
Let's identify which countries have reduced their CO2 emissions the most between 1995 and 2022:

```{r}
# Get CO2 emissions for all countries in 1995 and 2022
co2_1995 <- exio |>
    filter(year == 1995,
           matrix == "F_satellite",
           stressor %like% "%CO2%") |>
    group_by(region) |>
    summarise(co2_1995 = sum(value, na.rm = TRUE)) |>
    collect()

co2_2022 <- exio |>
    filter(year == 2022,
           matrix == "F_satellite",
           stressor %like% "%CO2%") |>
    group_by(region) |>
    summarise(co2_2022 = sum(value, na.rm = TRUE)) |>
    collect()

# Join the two datasets and calculate change
co2_change <- co2_1995 |>
    inner_join(co2_2022, by = "region") |>
    mutate(
        absolute_change = co2_2022 - co2_1995,
        percent_change = ((co2_2022 - co2_1995) / co2_1995) * 100
    ) |>
    filter(!grepl("^W", region)) |>  # Exclude rest-of-world regions (WA, WE, WF, WL, WM)
    arrange(absolute_change)

# View countries with the largest absolute decreases
head(co2_change, 5)
```

Now let's visualize the top 5 countries with the largest absolute reductions:

```{r}
# Get top 5 reducers
top_reducers <- co2_change |>
    head(5)

# Create a bar plot
ggplot(top_reducers, aes(x = reorder(region, absolute_change), y = absolute_change, fill = region)) +
    geom_col() +
    coord_flip() +
    scale_fill_brewer(palette = "Set2") +
    labs(
        title = "Top 5 Countries with Largest CO2 Emission Reductions",
        subtitle = "Change from 1995 to 2022",
        x = "Country",
        y = "Absolute Change in CO2 Emissions"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(face = "bold", size = 14),
        legend.position = "none"
    )
```

Let's also look at percentage reductions (excluding countries with very low baseline emissions):

```{r}
# Filter to countries with substantial 1995 emissions (>1000 units) to avoid outliers
substantial_emitters <- co2_change |>
    filter(co2_1995 > 1000) |>
    arrange(percent_change) |>
    head(5)

substantial_emitters

# Visualize percentage reductions
ggplot(substantial_emitters, aes(x = reorder(region, percent_change), y = percent_change, fill = region)) +
    geom_col() +
    coord_flip() +
    scale_fill_brewer(palette = "Dark2") +
    labs(
        title = "Top 5 Countries with Largest % CO2 Emission Reductions",
        subtitle = "Change from 1995 to 2022 (excluding very small emitters)",
        x = "Country",
        y = "Percent Change in CO2 Emissions (%)"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(face = "bold", size = 13),
        legend.position = "none"
    )
```
