---
title: "Module 1: Tabular Data"
subtitle: "Working with larger-than-RAM data using duckdbfs"
author: "ESPM 288"
format: 
    html:
        embed-resources: true
---

## Introduction

In this module, we will explore high-performance workflows for tabular data. We will use `duckdbfs` to work with datasets that are larger than available RAM by leveraging DuckDB's streaming and remote file access capabilities.

## Case Study: Global Supply Chains

We will be working with [EXIOBASE 3.8.1](https://source.coop/youssef-harby/exiobase-3), a global Multi-Regional Input-Output (MRIO) database. This dataset tracks economic transactions between sectors and regions, along with their environmental impacts (emissions, resource use, etc.).

**Data description:**
- **Coverage**: 44 countries + 5 rest-of-world regions.
- **Timeframe**: 1995–2022.
- **Content**: Economic transactions (Z matrix), final demand (Y matrix), and environmental stressors (F matrix).
- **Format**: Cloud-optimized Parquet, partitioned by year and matrix type.

## Setup

```{r}
library(duckdbfs)
library(dplyr)
library(ggplot2)
library(quarto)
```

## Exercise 1: connecting to remote data

We can open the entire dataset without downloading it using `open_dataset()`. The data is hosted on Source Cooperative. The `**` pattern allows recursive scanning of the partitioned parquet files.

```{r}
# Remote S3 path to EXIOBASE 3 (Source Cooperative)

duckdbfs::duckdb_secrets(
    key = "",
    secret = "",
    endpoint = "s3.amazonaws.com",
    region = "us-west-2"
)

# Open the entire dataset with ** pattern for recursive scanning
s3_url <- "s3://us-west-2.opendata.source.coop/youssef-harby/exiobase-3/4588235/parquet/**"

# Open the dataset lazily
exio <- open_dataset(s3_url)

# View the schema (column names and types) without reading data
glimpse(exio)
```

## Exercise 2: Efficient Filtering

The dataset is large. We should filter *before* collecting any data into R.

```{r}
exio |>
    filter(year == 2022, region == "US") |>
    head() |> # view the first 6 rows
    collect()
```

> **Task**: Construct a query to find the top 5 sectors in the US by CO2 emissions in 2022. Remember to check the column names in `exio` to find the appropriate emissions flow.

```{r}
# Solution: Top 5 sectors in US by CO2 emissions (2022)

# Target the F_satellite matrix directly to access stressor column
exio_f_sat <- open_dataset(
    "s3://us-west-2.opendata.source.coop/youssef-harby/exiobase-3/4588235/parquet/year=2022/format=ixi/matrix=F_satellite/"
)

exio_f_sat |>
    filter(
        grepl("CO2", stressor)
    ) |>
    group_by(sector) |>
    summarize(total_co2 = sum(value, na.rm = TRUE), .groups = "drop") |>
    arrange(desc(total_co2)) |>
    head(5) |>
    collect()
```

## Exercise 3: CO2 Emissions Over Time by Region (Top 5 Emitters)

First, let's identify the top 5 CO2 emitting regions in 2022:

```{r}
# Find the top 5 emitting regions in 2022
top_emitters_2022 <- exio |>
    filter(year == 2022,
           matrix == "F_satellite",
           stressor %like% "%CO2%") |>
    group_by(region) |>
    summarise(total_co2 = sum(value, na.rm = TRUE)) |>
    arrange(desc(total_co2)) |>
    head(5) |>
    collect()

top_emitters_2022
```

Now, let's get the time series data for these top 5 regions across all years:

```{r}
# Get the list of top 5 regions
top_5_regions <- top_emitters_2022$region

# Query CO2 emissions for these regions across all years
co2_timeseries <- exio |>
    filter(matrix == "F_satellite",
           stressor %like% "%CO2%",
           region %in% top_5_regions) |>
    group_by(year, region) |>
    summarise(total_co2 = sum(value, na.rm = TRUE)) |>
    arrange(year, region) |>
    collect()

# View the data
head(co2_timeseries, 20)
```

Create a line plot showing CO2 emissions over time for the top 5 regions:

```{r}
# Create a mapping of region codes to full names
region_names <- c(
    "CN" = "China",
    "US" = "United States",
    "IN" = "India",
    "RU" = "Russia",
    "JP" = "Japan",
    "DE" = "Germany",
    "GB" = "United Kingdom",
    "KR" = "South Korea",
    "CA" = "Canada",
    "BR" = "Brazil",
    "WE" = "Rest of Europe",
    "WA" = "Rest of Asia",
    "WF" = "Rest of Africa",
    "WL" = "Rest of Latin America",
    "WM" = "Rest of Middle East"
)

# Add full region names to the data
co2_timeseries <- co2_timeseries |>
    mutate(region_name = ifelse(region %in% names(region_names), 
                                 region_names[region], 
                                 region))

# Create the time series plot
co2_plot <- ggplot(co2_timeseries, aes(x = year, y = total_co2, color = region_name)) +
    geom_line(linewidth = 1.2) +
    geom_point(size = 2.5) +
    labs(
        title = "CO2 Emissions Over Time: Top 5 Emitting Regions",
        subtitle = "Based on EXIOBASE 3.8.1 data (1995-2022)",
        x = "Year",
        y = "Total CO2 Emissions",
        color = "Region"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(face = "bold", size = 14),
        legend.position = "right",
        legend.title = element_text(face = "bold", size = 11),
        legend.text = element_text(size = 10),
        legend.background = element_rect(fill = "white", color = "gray80"),
        legend.key.size = unit(1, "cm")
    ) +
    scale_x_continuous(breaks = seq(1995, 2022, by = 5))

co2_plot

# Save the plot
ggsave("co2_regions_timeseries.png", co2_plot, width = 10, height = 6, dpi = 300)
```

## Exercise 4: Regions with Largest CO2 Emission Reductions

Let's identify which regions have reduced their CO2 emissions the most between 1995 and 2022:

```{r}
# Get CO2 emissions for all regions in 1995 and 2022
co2_1995 <- exio |>
    filter(year == 1995,
           matrix == "F_satellite",
           stressor %like% "%CO2%") |>
    group_by(region) |>
    summarise(co2_1995 = sum(value, na.rm = TRUE)) |>
    collect()

co2_2022 <- exio |>
    filter(year == 2022,
           matrix == "F_satellite",
           stressor %like% "%CO2%") |>
    group_by(region) |>
    summarise(co2_2022 = sum(value, na.rm = TRUE)) |>
    collect()

# Join the two datasets and calculate change
co2_change <- co2_1995 |>
    inner_join(co2_2022, by = "region") |>
    mutate(
        absolute_change = co2_2022 - co2_1995,
        percent_change = ((co2_2022 - co2_1995) / co2_1995) * 100
    ) |>
    filter(grepl("^W", region)) |>  # Include only rest-of-world regions (WA, WE, WF, WL, WM)
    arrange(absolute_change)

# View regions with the largest absolute decreases
head(co2_change, 5)
```

Now let's visualize the top 5 regions with the largest absolute reductions:

```{r}
# Get top 5 reducers
top_reducers <- co2_change |>
    head(5)

# Create a bar plot
reduction_plot <- ggplot(top_reducers, aes(x = reorder(region, absolute_change), y = absolute_change, fill = region)) +
    geom_col() +
    coord_flip() +
    scale_fill_brewer(palette = "Set2") +
    labs(
        title = "Top 5 Regions with Largest CO2 Emission Reductions",
        subtitle = "Change from 1995 to 2022",
        x = "Region",
        y = "Absolute Change in CO2 Emissions"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(face = "bold", size = 14),
        legend.position = "none"
    )

reduction_plot

# Save the plot
ggsave("co2_absolute_reduction.png", reduction_plot, width = 10, height = 6, dpi = 300)
```

Let's also look at percentage reductions (excluding countries with very low baseline emissions):

```{r}
# Filter to countries with substantial 1995 emissions (>1000 units) to avoid outliers
substantial_emitters <- co2_change |>
    filter(co2_1995 > 1000) |>
    arrange(percent_change) |>
    head(5)

substantial_emitters

# Visualize percentage reductions
percent_plot <- ggplot(substantial_emitters, aes(x = reorder(region, percent_change), y = percent_change, fill = region)) +
    geom_col() +
    coord_flip() +
    scale_fill_brewer(palette = "Dark2") +
    labs(
        title = "Top 5 Regions with Largest % CO2 Emission Reductions",
        subtitle = "Change from 1995 to 2022 (excluding very small emitters)",
        x = "Region",
        y = "Percent Change in CO2 Emissions (%)"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(face = "bold", size = 13),
        legend.position = "none"
    )

percent_plot

# Save the plot
ggsave("co2_percent_reduction.png", percent_plot, width = 10, height = 6, dpi = 300)
```

## Exercise 5: CO2 Emissions Over Time for Top 5 Emitting Countries

Now let's look at individual countries (excluding rest-of-world aggregates) to see the top 5 emitting countries:

```{r}
# Find the top 5 emitting countries in 2022 (excluding W regions)
top_countries_2022 <- exio |>
    filter(year == 2022,
           matrix == "F_satellite",
           stressor %like% "%CO2%") |>
    group_by(region) |>
    summarise(total_co2 = sum(value, na.rm = TRUE)) |>
    filter(!grepl("^W", region)) |>  # Exclude rest-of-world regions
    arrange(desc(total_co2)) |>
    head(5) |>
    collect()

top_countries_2022
```

Get time series data for these top 5 countries across all years:

```{r}
# Get the list of top 5 countries
top_5_countries <- top_countries_2022$region

# Query CO2 emissions for these countries across all years
co2_countries_timeseries <- exio |>
    filter(matrix == "F_satellite",
           stressor %like% "%CO2%",
           region %in% top_5_countries) |>
    group_by(year, region) |>
    summarise(total_co2 = sum(value, na.rm = TRUE)) |>
    arrange(year, region) |>
    collect()

# View the data
head(co2_countries_timeseries, 20)
```

Create a line plot showing CO2 emissions over time for the top 5 countries:

```{r}
# Add full country names to the data
co2_countries_timeseries <- co2_countries_timeseries |>
    mutate(country_name = ifelse(region %in% names(region_names), 
                                  region_names[region], 
                                  region))

# Create the time series plot
countries_plot <- ggplot(co2_countries_timeseries, aes(x = year, y = total_co2, color = country_name)) +
    geom_line(linewidth = 1.2) +
    geom_point(size = 2.5) +
    labs(
        title = "CO2 Emissions Over Time: Top 5 Emitting Countries",
        subtitle = "Based on EXIOBASE 3.8.1 data (1995-2022)",
        x = "Year",
        y = "Total CO2 Emissions",
        color = "Country"
    ) +
    theme_minimal() +
    theme(
        plot.title = element_text(face = "bold", size = 14),
        legend.position = "right",
        legend.title = element_text(face = "bold", size = 11),
        legend.text = element_text(size = 10),
        legend.background = element_rect(fill = "white", color = "gray80"),
        legend.key.size = unit(1, "cm")
    ) +
    scale_x_continuous(breaks = seq(1995, 2022, by = 5))

countries_plot

# Save the plot
ggsave("co2_countries_timeseries.png", countries_plot, width = 10, height = 6, dpi = 300)
```
